Scheduling Musical Events in Max/MSP with Scheme For Max

### 2021-08-09 paper-1/paper-1.tex
copy edits from reviewers

In both the fields of live-coding and algorithmic music, the ability for
the performer/composer to schedule events in the future with high temporal
accuracy is of major benefit. While this is possible in the Max visual
patching language, it is cumbersome and thus the implementation is not
optimal for live-coding or algorithmic music where the speed with which
the user can accomplish this is an important consideration (i.e., fast
enough not to bore an audience or to be frustrating while composing).

@@

Scheme For Max (a.k.a. S4M) is an open-source external developed by the 
author that embeds the s7 Scheme interpreter, a Scheme dialect created by
Bill Schottstaedt at the  Center for Computer Research in Music and Acoustics
(CCRMA) at Stanford University. Originally developed
from TinyScheme, s7 is a Scheme dialect designed for use in computer music
platforms and is used in various music programs such as the Snd editor and the
Common Music algorithmic composition platform \cite{Schottstaedt2021}.
Similar in functionality to the built
in \textbf{js} object, the \textbf{s4m} object enables the user to program Max in s7 Scheme,
and provides Scheme functions to interact with the Max engine and environment through
a foreign function interface implemented in C using the Max C SDK and API.

The author believes that S4M extends Max in a way that is of significant value to 
the algorithmic musician and live-coding performer, as well as more generally to
to the broader Max community. The ability to update
a running Scheme program during playback is a major benefit
and is a capability that has been, while technically possible, impractical
with existing solutions such as JavaScript. Additionally, the syntax
of Scheme bears a convenient similarity to Max message syntax, and thus 
one can easily use Max visual widgets and messages to generate small Scheme programs in the
patcher. Potentially the most interesting benefit is the ease with which the user 
can create and schedule functions for evaluation in the future, with 
input into these functions coming from Max patching widgets, and flexible control
over the current and future evaluation context of the functions and variables used.

@@

In addition to this event-based message execution model, also called ``control messages", Max 
supports a stream-based digital 
audio execution model, originally provided separately as a product called ``MSP", but now included 
as part of the single Max product.
MSP activity normally runs in a separate thread from the event/message Max operations, and uses a  
a separate class of objects that pass constant streams of digital audio to each other through differently 
colored patch cords (though MSP objects may additionally receive control messages for controlling parameters). As  
S4M executes only at the event/message level and does not implement DSP operations, MSP is not 
discussed further here.

@@

Internally, Max messages are data entities consisting of a symbol that acts
as a Smalltalk-style message selector and an optional array of Max atoms. 
Each atom entity contains a member for the atom type and another for its value, 
where the type may be any of \textbf{int,} \textbf{float,} or \textbf{symbol}. 
Note that while the message selector symbol is always present at the C level,
in the visual patcher the selector may be implicit and hidden from the user. 
(i.e., the message originating from a message box that appears to contain only an integer
will actually consist of the selector ``int" followed by an atom storing the numerical value.)
The symbol ``bang" is a special symbol that can be used for a one-element message 
that essentially means ``run", and the act of triggering execution by sending a bang
message from the \textbf{bang} object is called ``banging" in the nomenclature.
Taken together, this means that in Max there are five kinds of messages: 
\textbf{bang}, \textbf{symbol}, \textbf{int}, \textbf{float}, and \textbf{list}. 
(The bang message is technically a symbol message
of ``bang", but this is essentially treated as a type of its own in the nomenclature.) 
\cite{Puckette2002}

@@

As an example, in the patcher screenshot above we see a \textbf{message-box} object that will update
its state (and visual display) with the symbols ``hello" and ``world" when sent the message 
\textbf{set hello world}. Near it is a circular \textbf{bang} object, which when clicked will send the
\textbf{message-box} the \textbf{bang} message, causing it to output the message \textbf{list hello world}.
This message will be received by the \textbf{zl.len} object, which counts the number of elements 
in any lists it receives, immediately outputting the count. The three columns in the screen-shot
show the patch as it is prior to any clicks, as it is after clicking \textbf{message-box} 1,
and as it is after clicking both \textbf{message-box} 1 and the \textbf{bang}.

@@

The second facility for dynamic code evaluation in S4M is more unusual.
In a Max external, there exists a binding option to catch any previously
uncaught messages, so that a message with an unrecognized selector 
can be handled by a generic dispatching method. In S4M, this is used to 
catch any message that is not already reserved, with any additional 
arguments passed in as a Max list of atoms.  The \textbf{s4m} object interprets
the message to mean \textit{``evaluate this selector plus its additional atoms 
as if it is a Scheme list enclosed by parentheses"}.
Because both Max and Scheme uses white-space as separators and very little
in the way of diacritical syntax, this enables a wide variety
of simple Scheme calls to be made with little ceremony. The
individual atoms of the Max message are converted to Scheme tokens,
assembled into a list, and this list is evaluated. While seemingly simple,
this facility is of significant utility to the Max programmer, as combined
with the Max message interpolation facility, it enables the user to very
easily assemble Max widgets that trigger Scheme functions. Most anything that
can be expressed in a single s-expression without inner nesting can be used.
In figure 4, we see a \textbf{number-box} outputting to a \textbf{message-box} using Max's 
variable interpolation (the \textbf{\$1}). 
Clicking the \textbf{bang} above the \textbf{number-box} or changing the number in 
the box will result in a message of 
\textbf{set-volume X} being sent to the \textbf{s4m} object, which will then call
into Scheme with the code \textbf{(set-volume X)}.

@@

One of the major advantages of Max compared to other computer music platforms 
is the ease with which the user can create interactive environments in which the performer's actions 
change musical parameters in real-time in complex ways, enabling 
musically rich performances. It is, for example, trivial to add GUI elements to change
musical parameters, and only slightly less trivial to add handlers for MIDI input, 
so that users can connect to their programs physical devices such as
piano-style keyboards, and mixing board knobs, faders, and buttons.

@@

Let us imagine instead that the \textbf{bang} message should delay the activity
(posting to the console) by 1000 ms. We can use the Max \textbf{clock} facility for high-accuracy
delay as it accepts floats for the ms time value. (Note that the actual delay time
will be offset to the nearest signal vector boundary, determined by the
users ``Signal Vector Size" audio setting, but subsequent clock calls 
adjust for this, maintaining long term temporal accuracy at sample rate resolution.)

@@

\begin{itemize}
\item A \textbf{t\_s4m\_clock\_callback} data structure is dynamically allocated and used to store a
  reference to the \textbf{s4m} object and to the symbolic handle passed from Scheme.
\item A Max \textbf{clock} is created, passing in a void pointer to clock callback structure
  and binding the clock to a generic clock callback function in C. The clock's timer is started
\item The clock callback struct is also stored in a C hash-table (Max's hashtab implementation),
  keyed by the handle, so that cancellation or object deletion can use this to find all clocks.
\item When the generic clock callback function runs (time has elapsed), it uses the pointer argument to get 
  the delay handle and the instantiated \textbf{s4m} object, through which it 
  can also get the s7 interpreter pointer.
\item The s7 interpreter and handle are then used to call the Scheme function
  \textbf{s4m-execute-callback,} which will run our delayed Scheme function as previously explained.
\item The generic C callback then removes the clock reference from the C hash-table,
  deletes the clock, and frees the memory allocated for the callback struct.
\end{itemize}

@@

The scheduling system describe does have several limitations worth noting.
The s7 Scheme implementation includes a tracing garbage collector 
that must run occasionally, pausing other processing \cite{Matheussen2020}.
The garbage collector can, however, be paused and run on demand. Tests on the author's
system, with the garbage collector forced to run every 100 ms to prevent
significant build up, showed the garbage collection taking an average of 1 to 2 ms per pass
when run in a substantial application (a 16 track sequencer written entirely in S4M
with the computer also generating audio for the tracks with commercial
software synthesizers).  
The audio rendered was recorded and examined
examined in a digital audio workstation to check for timing discrepancies. The
results of these preliminary tests were that timing was unaffected by the 
garbage collector so long as Max was configured with 
a large enough output buffer to allow the collector to run within the 
latency period of this buffer.
Using Max with an output buffer of 256 samples produces an output latency
of approximately 6-8 ms (depending on the Max signal vector as well), which
was large enough for the collector to run without issue, but is 
low enough to be acceptable for real time performance. With software synthesis
loads at closer to the maximum possible on the author's computer,
an output buffer of 512 samples (approx. 11 ms output latency) was necessary to prevent
any timing errors. This is encouraging as these buffer sizes are not atypical
for commercial music production on digital audio workstations and are widely
considered as being in acceptable ranges for real-time music systems \cite{Brandt1998}.

In addition to output latency, there is the factor of internal latency.
Max, like many computer music programming environments, generates audio 
in vectors of samples, with
each vector generated from one processing pass, and (normally) control
rate passes conducted also a maximum of once per audio vector pass \cite{Puckette2002}.
The size of this signal vector thus has a large effect on performance, with larger
vectors requiring significantly less CPU use. As audio rendering of note onsets originating
from control messages (the normal scenario) must thus be aligned with 
signal vector boundaries, this produces delays of note onsets by up to
the time period of one vector, a phenomenon known as ``timing jitter" in the nomenclature.
However, the more recent scheduling facilities in Max, including the 
clock facilities used by S4M, are implemented to compensate for this jitter,
maintaining long term temporal accuracy despite note onset delays. The mechanism
by which this is done is unfortunately not publicly documented or available
in open-source code, but one can assume that it is some variant of the common pattern 
whereby scheduled events are aware of when they are supposed to be running
(``logical-time"), and if they in turn schedule subsequent events, the time for these subsequent
events will be shifted to compensate for the jitter, based 
on the logical time of the scheduling event, rather than its
actual (post-jitter) real time \cite{Anderson1986}.
The author also conducted personal tests of 
this compensation, comparing both high and low signal vector sizes to produce
variable delays in note onsets, and it was determined
that so long as the overall output latency is again large enough to allow
the GC to run in time, self-scheduling events in a long piece do stay
accurate in that jitter does not accumulate over the piece - 
note onsets stay accurate to within the delay of one signal vector of samples,
regardless of the length of the piece of music.

@@

\section{Conclusion}
The Scheme For Max project brings a new way of working 
to the Max platform, especially with regard to flexible and accurate 
scheduling of future events.  The author believes this will be
of significant benefit to those working in the fields of algorithmic
music and live coding, and will also be useful to the more general Max 
user base as a way to script Max in a high level language that runs in
the high-priority thread. In the author's experience, 
working in Scheme within Max has been highly productive, with the
language well suited both to expressing musical concepts and to working
with the Max platform. The ability to update code during play-back
has been an especially productive workflow for building complex 
sequencing and algorithmic composition tools.

@@

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
%%\begin{acks}
%%Scheme For Max would not exist without the work of many people. 
%%Bill Schottstaedt created Common Lisp Music and the s7 Scheme implementation,
%%and has been invaluable as a source of advice on s7. Heinrich Taube
%%created Common Music and has been an inspiration through his seminal book
%%on algorithmic composition with Lisp, "Notes from the Metalevel".
%%Luigi Castelli and Alex Harker helped with programming questions
%%and build assistance. 
%%S4M of course depends on Max and thus owes a huge debt to 
%%Miller Puckette and David Zicarelli. And finally, Andrew Schloss has
%%been a source of encouragement and support through the years of 
%%exploration that led to the Scheme For Max project.
%%
%%\end{acks}

>>>>

In both the fields of live-coding and algorithmic music, the ability for
the performer and composer to schedule events in the future with high temporal
accuracy is of major benefit. While this is possible in the Max visual
patching language, it is cumbersome and thus the implementation is not
optimal for live-coding or algorithmic music where the speed with which
the user can accomplish this is an important consideration (i.e., fast
enough not to bore an audience or to be frustrating while composing).

@@

Scheme For Max (a.k.a. S4M) is an open-source external developed by the 
author that embeds the s7 Scheme interpreter, a Scheme dialect created by
Bill Schottstaedt at the  Center for Computer Research in Music and Acoustics
(CCRMA) at Stanford University. Originally developed
from TinyScheme, s7 is a Scheme dialect designed for use in computer music
platforms and is used in various music programs such as the Snd editor and the
Common Music algorithmic composition platform \cite{Schottstaedt2021}.
Similar in functionality to the built-in \textbf{js} object, 
the \textbf{s4m} object enables the user to program Max in s7 Scheme,
and provides Scheme functions to interact with the Max engine and environment through
a foreign function interface implemented in C using the Max C SDK and API.

The author believes that S4M extends Max in a way that is of significant value to 
the algorithmic musician and live-coding performer, as well as more generally 
to the broader Max community. The ability to update
a running Scheme program during playback is a major benefit
and is a capability that has been, while technically possible, impractical
with existing solutions such as JavaScript. Additionally, the syntax
of Scheme bears a convenient similarity to Max message syntax, and thus 
one can easily use Max visual widgets and messages to generate small Scheme programs in the
patcher. Potentially the most interesting benefit is the ease with which the user 
can create and schedule functions for evaluation in the future, with 
input into these functions coming from Max patching widgets, and flexible control
over the current and future evaluation context of the functions and variables used.

@@

In addition to this event-based message execution model, also called ``control messages", Max 
supports a stream-based digital 
audio execution model, originally provided separately as a product called ``MSP", but now included 
as part of the single Max product.
MSP activity normally runs in a separate thread from the event/message Max operations, and uses 
a separate class of objects that pass constant streams of digital audio to each other through differently 
colored patch cords (though MSP objects may additionally receive control messages for controlling parameters). As  
S4M executes only at the event/message level and does not implement DSP operations, MSP is not 
discussed further here.

@@

Internally, Max messages are data entities consisting of a symbol that acts
as a Smalltalk-style message selector and an optional array of Max atoms. 
Each atom entity contains a member for the atom type and another for its value, 
where the type may be any of \textbf{int,} \textbf{float,} or \textbf{symbol}. 
Note that while the message selector symbol is always present at the C level,
in the visual patcher the selector may be implicit and hidden from the user. 
(i.e., the message originating from a message box that appears to contain only an integer
will actually consist of the selector ``int" followed by an atom storing the numerical value.)
The symbol ``bang" is a special symbol that can be used for a one-element message 
that essentially means ``run", and the act of triggering execution by sending a bang
message from the \textbf{bang} object is called ``banging" in the nomenclature.
(Note that there are both messages and objects called ``bang"; the \textbf{bang} 
object is an object that sends a bang message on receipt of any message.)
Taken together, this means that in Max there are five kinds of messages: 
\textbf{bang}, \textbf{symbol}, \textbf{int}, \textbf{float}, and \textbf{list}. 
\cite{Puckette2002}
(The bang message is technically a symbol message
of ``bang", but this is essentially treated as a type of its own in the nomenclature.)

@@

As an example, in the patcher screenshot in figure 3 we see a \textbf{message-box} object that will update
its state (and visual display) with the symbols ``hello" and ``world" when sent the message 
\textbf{set hello world}. Near it is a circular \textbf{bang} object, which when clicked will send the
\textbf{message-box} the \textbf{bang} message, causing it to output the message \textbf{list hello world}.
This message will be received by the \textbf{zl.len} object, which counts the number of elements 
in any lists it receives, immediately outputting the count. The three columns in the screen-shot
show the patch as it is prior to any clicks, as it is after clicking \textbf{message-box} 1,
and as it is after clicking both \textbf{message-box} 1 and the \textbf{bang}.

@@

The second facility for dynamic code evaluation in S4M is more unusual.
In a Max external, there exists a binding option to catch any previously
uncaught messages, so that a message with an unrecognized selector 
can be handled by a generic dispatching method. In S4M, this is used to 
catch any message that is not already reserved, with any additional 
arguments passed in as a Max list of atoms.  The \textbf{s4m} object interprets
the message to mean \textit{``evaluate this selector plus its additional atoms 
as if it is a Scheme list enclosed by parentheses"}.
Because both Max and Scheme uses white-space as separators and very little
in the way of diacritical syntax, this enables a wide variety
of simple Scheme calls to be made with little ceremony. The
individual atoms of the Max message are converted to Scheme tokens,
assembled into a list, and this list is evaluated. While seemingly simple,
this facility is of significant utility to the Max programmer, as combined
with the Max message interpolation facility, it enables the user to very
easily assemble Max widgets that trigger Scheme functions. Most anything that
can be expressed in a single s-expression without inner nesting can be used.
In figure 5, we see a \textbf{number-box} outputting to a \textbf{message-box} using Max's 
variable interpolation (the \textbf{\$1}). 
Clicking the \textbf{bang} above the \textbf{number-box} or changing the number in 
the box will result in a message of 
\textbf{set-volume X} being sent to the \textbf{s4m} object, which will then call
into Scheme with the code \textbf{(set-volume X)}.

@@

One of the major advantages of Max compared to other computer music platforms 
is the ease with which the user can create interactive environments in which the performer's actions 
change musical parameters in complex ways, enabling 
musically rich performances. It is, for example, trivial to add GUI elements to change
musical parameters, and only slightly less trivial to add handlers for MIDI input, 
so that users can connect to their programs physical devices such as
piano-style keyboards, and mixing board knobs, faders, and buttons.

@@

Let us imagine instead that the \textbf{bang} message should delay the activity
(posting to the console) by 1000 ms. We can use the Max \textbf{clock} facility for high-accuracy
delay as it accepts floats for the ms time value. (Note that the actual delay time
will be offset to the nearest signal vector boundary, determined by the
users ``Signal Vector Size" audio setting, but subsequent clock calls 
adjust for this, maintaining long-term temporal accuracy at sample rate resolution.)

@@

\begin{itemize}
\item A \textbf{t\_s4m\_clock\_callback} data structure is dynamically allocated and used to store a
  reference to the \textbf{s4m} object and to the symbolic handle passed from Scheme.
\item A Max \textbf{clock} is created, passing in a void pointer to clock callback structure
  and binding the clock to a generic clock callback function in C. The clock's timer is started.
\item The clock callback struct is also stored in a C hash-table (Max's hashtab implementation),
  keyed by the handle, so that cancellation or object deletion can use this to find all clocks.
\item When the generic clock callback function runs (time has elapsed), it uses the pointer argument to get 
  the delay handle and the instantiated \textbf{s4m} object, through which it 
  can also get the s7 interpreter pointer.
\item The s7 interpreter and handle are then used to call the Scheme function
  \textbf{s4m-execute-callback,} which will run our delayed Scheme function as previously explained.
\item The generic C callback then removes the clock reference from the C hash-table,
  deletes the clock, and frees the memory allocated for the callback struct.
\end{itemize}

@@

The scheduling system describe does have several limitations worth noting.
The s7 Scheme implementation includes a tracing garbage collector 
that must run occasionally, pausing other processing \cite{Matheussen2020}.
The garbage collector can, however, be paused and run on demand. Tests on the author's
system, with the garbage collector forced to run every 100 ms to prevent
significant build up, showed the garbage collection taking an average of 1 to 2 ms per pass
when run in a substantial application (a 16 track sequencer written entirely in S4M
with the computer also generating audio for the tracks with commercial
software synthesizers).  
The audio rendered was recorded and examined
in a digital audio workstation to check for timing discrepancies. The
results of these preliminary tests were that timing was unaffected by the 
garbage collector so long as Max was configured with 
a large enough output buffer to allow the collector to run within the 
latency period of this buffer.
Using Max with an output buffer of 256 samples produces an output latency
of approximately 6-8 ms (depending on the Max signal vector as well), which
was large enough for the collector to run without issue, but is 
low enough to be acceptable for real-time performance. With software synthesis
loads at closer to the maximum possible on the author's computer,
an output buffer of 512 samples (approx. 11 ms output latency) was necessary to prevent
any timing errors. This is encouraging as these buffer sizes are not atypical
for commercial music production on digital audio workstations and are widely
considered as being in acceptable ranges for real-time music systems \cite{Brandt1998}.

In addition to output latency, there is the factor of internal latency.
Max, like many computer music programming environments, generates audio 
in vectors of samples, with
each vector generated from one processing pass, and (normally) control
rate passes conducted also a maximum of once per audio vector pass \cite{Puckette2002}.
The size of this signal vector thus has a large effect on performance, with larger
vectors requiring significantly less CPU use. As audio rendering of note onsets originating
from control messages (the normal scenario) must thus be aligned with 
signal vector boundaries, this produces delays of note onsets by up to
the time period of one vector, a phenomenon known as ``timing jitter" in the nomenclature.
However, the more recent scheduling facilities in Max, including the 
clock facilities used by S4M, are implemented to compensate for this jitter,
maintaining long-term temporal accuracy despite note onset delays. The mechanism
by which this is done is unfortunately not publicly documented or available
in open-source code, but one can assume that it is some variant of the common pattern 
whereby scheduled events are aware of when they are supposed to be running
(``logical-time"), and if they in turn schedule subsequent events, the time for these subsequent
events will be shifted to compensate for the jitter, based 
on the logical time of the scheduling event, rather than its
actual (post-jitter) real time \cite{Anderson1986}.
The author also conducted personal tests of 
this compensation, comparing both high and low signal vector sizes to produce
variable delays in note onsets, and it was determined
that so long as the overall output latency is again large enough to allow
the GC to run in time, self-scheduling events in a long piece do stay
accurate in that jitter does not accumulate over the piece - 
note onsets stay accurate to within the delay of one signal vector of samples,
regardless of the length of the piece of music.

@@

\section{Conclusion}
The Scheme For Max project brings a new way of working 
to the Max platform, especially with regard to flexible and accurate 
scheduling of future events.  The author believes this will be
of significant benefit to those working in the fields of algorithmic
music and live coding, and will also be useful to the more general Max 
user base as a way to script Max in a high-level language that runs in
the high-priority thread. In the author's experience, 
working in Scheme within Max has been highly productive, with the
language well suited both to expressing musical concepts and to working
with the Max platform. The ability to update code during play-back
has been an especially productive workflow for building complex 
sequencing and algorithmic composition tools.

@@

https://github.com//scheme-for-max
%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
%%\begin{acks}
%%Scheme For Max would not exist without the work of many people. 
%%Bill Schottstaedt created Common Lisp Music and the s7 Scheme implementation,
%%and has been invaluable as a source of advice on s7. Heinrich Taube
%%created Common Music and has been an inspiration through his seminal book
%%on algorithmic composition with Lisp, "Notes from the Metalevel".
%%Luigi Castelli and Alex Harker helped with programming questions
%%and build assistance. 
%%S4M of course depends on Max and thus owes a huge debt to 
%%Miller Puckette and David Zicarelli. And finally, Andrew Schloss has
%%been a source of encouragement and support through the years of 
%%exploration that led to the Scheme For Max project.
%%
%%\end{acks}
